Flaring star-planet interactions in star-planet systems in Kepler and TESS

### 2023-05-11 src/tex/ms.tex
[showyourwork] overleaf sync

\subsection{Flares}
\label{sec:intro:flares}
In this study, we will look at stellar flares as another way of dissipating the energy deposited in the magnetic field. Flares are strong and impulsive eruptions released by reconnection and subsequent relaxation of field lines in the corona~\citep{svestka1976solar,priest2002magnetic}. In contrast to solar flares, stellar flares often are extremely energetic, sometimes enhancing the stellar flux by orders of magnitude~\citep{maehara2012superflares, shibayama2013superflares, paudel2018k2} so that they can be detected in stars over 1kpc away~\citep{chang2015}. But since they occur randomly in time, individual events can be hard to catch. %We can observe flares in nearly all electromagnetic bands -- from radio to X-ray~\citep{benz2010physical}.

@@

There could also be physical reasons for increased flaring interaction-in systems with close spin-orbit commensurability. \citet{lanza2022model} suggest that tidal torque of the planet on the star excites resonant oscillations in the interior stellar magnetic field, which in turn leads to a resonance between stellar spin and planetary orbit. These oscillations can lead to a surface pattern of emerging magnetic field that rotates with the spin-orbit commensurable period. Systems with spin-orbit resonance would then show clearer interaction signal, because the resonance would stabilize the pattern of regions with strong magnetic field that favor planet-induced flaring. The stable active regions might withstand the diluting effect of differential rotation~(see Section~\ref{sec:discussion:rotsyn}). However, the \citet{lanza2022model} requires tidal interaction, which may be too weak in systems like AU Mic, where the planet is too small and distant for the proposed effect, but that still show spin-orbit commensurability and signs of flaring star-planet interaction.

@@

\subsection{Tidal interaction}
\begin{figure*}[t]
    \script{paper_tidal_interaction.py}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/PAPER_tidal_AD_test.png}
        \caption{
           \textbf{Left panel}: Blue color indicates that the star rotates slower than the planet orbits, so that angular momentum is transferred from the orbit to the spin of the star. Grey color is the reverse. \textbf{Middle panel}: \textbf{Right panel}: 
        }
        \label{fig:tidal}
    \end{centering}
\end{figure*}

\begin{table*}
\footnotesize
\movetableright=-20mm
\script{paper_table_tidal.py}
\caption{Flaring tidal star-planet interaction.}
\input{output/table_tidal.tex}
    \label{tab:maintable_der}
 \tablerefs{\input{output/table_tidal_bibstring.tex}}
\end{table*}

Tidal interaction can also lead to enhanced activity, globally, and locally. Locally, we expect to see periodicity with half the orbital period of the planet, corresponding to the two tidal bulges forming in the stellar envelope. XXXX tidal theory here XXXX

@@

We applied the same technique as described in Section~\ref{sec:methods:adtest} to determine the significance of a local tidal SPI signal. Following~\citet{ilic2022tidal}, we compare the resulting $p$-values with three different models of tidal interaction as proxies for expected excess flaring due to tides in Fig.~\ref{fig:tidal}. We list the derived values in Table~(XXXX).

The first model measures the torque $\partial L / \partial t$, exerted by angular momentum transfer from star to planet or from planet to star~(\citealt{penev2012constraining}, their Eqns. 1 and 2). We choose the tidal quality factor $Q_*=10^7$. Our results do not depend on the choice of $Q_*$, as we only need to know the tidal torque down to a proportionality factor. We also make the simplifying assumption that the rotation period of the convective envelope is the same as the surface rotation period measured as described in Section~\ref{sec:data:rotationperiods}, ignoring differential rotation effects. \textcolor{red}{Why doesn't the torque depend on the stellar rotation velocity, except for the sign of interaction?}.

The second model considers the tidal dissipation timescale -- the shorter the timescale, the higher the power of the interaction~\citep{zahn1977tidal}. Following~\citet{ilic2022tidal}, we adopt Eqn.2 from~\citep{albrecht2012obliquities} for the tidal dissipation timescale in the convective envelope. We assume that all stars in our sample have a convective envelope because they flare, and flaring requires a dynamo-generated magnetic field, which can only operate in stars with convective envelopes. Most of our stars are clearly below the Kraft break at spectral type F5~(XXXX), above which stars have a radiative envelope, except for KOI-12, which is an F5 sub-giant~\citep{frasca2016activity}.

The third model uses the gravitational perturbation as a proxy for the interaction~\citep{cuntz2000stellar}. All models require the planetary and stellar mass to be known which we collect from the literature. For missing planetary masses, radii are known, so we can use the mass-radius relations implemented in \texttt{forecaster}~\citep{chen2017probabilistic}. GJ 3323, GJ 674 and GJ 3082 were also missing stellar mass estimates, which we obtained from the \citep{mann2015how, mann2016erratum} relations between absolute Ks magnitude~\citep{skrutskie2006two} and stellar mass for M dwarfs using the distances from \citet{bailer-jones2021estimating}. We list the masses in Table~(XXXX).

>>>>

\subsection{Flares}
\label{sec:intro:flares}
In this study, we will look at stellar flares as another way of dissipating the energy deposited in the magnetic field. Flares are strong and impulsive eruptions released by reconnection and subsequent relaxation of field lines in the corona~\citep{svestka1976solar,priest2002magnetic}. In contrast to solar flares, stellar flares often are extremely energetic, sometimes enhancing the stellar flux by orders of magnitude~\citep{maehara2012superflares, shibayama2013superflares, paudel2018k2} so that they can be detected in stars over 1kpc away~\citep{chang2015photometric}. But since they occur randomly in time, individual events can be hard to catch. %We can observe flares in nearly all electromagnetic bands -- from radio to X-ray~\citep{benz2010physical}.

@@

There could also be physical reasons for increased flaring interaction-in systems with close spin-orbit commensurability. \citet{lanza2022model} suggest that tidal torque of the planet on the star excites resonant oscillations in the interior stellar magnetic field, which in turn leads to a resonance between stellar spin and planetary orbit. These oscillations can lead to a surface pattern of emerging magnetic field that rotates with the spin-orbit commensurable period. Systems with spin-orbit resonance would then show clearer interaction signal, because the resonance would stabilize the pattern of regions with strong magnetic field that favor planet-induced flaring. The stable active regions might withstand the diluting effect of differential rotation~(see Section~\ref{sec:discussion:rotsyn}). However, the \citet{lanza2022model} requires tidal interaction, which may be too weak in systems like AU Mic~(see also Section~\ref{sec:discussion:tidal}), where the planet is too small and distant for the proposed effect, but that still show spin-orbit commensurability and signs of flaring star-planet interaction.

@@

\subsection{Tidal interaction}
\label{sec:discussion:tidal}
\begin{figure*}[t]
    \script{paper_tidal_interaction.py}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/PAPER_tidal_AD_test.png}
        \caption{Deviation from random flare times in phase with $P_{orb}/2$, i.e., in phase with tidal bulges raised on the star, compared to different models of tidal interaction.
           \textbf{Left panel}: The tidal torque in the system for stars with a convective envelope. Blue color indicates that the star rotates slower than the planet orbits, so that angular momentum is transferred from the orbit to the spin of the star. Grey color is the reverse. \textbf{Middle panel}: The tidal dissipation timescale. \textbf{Right panel}:  Relative gravitational perturbation in the star.}
        \label{fig:tidal}
    \end{centering}
\end{figure*}

\begin{table*}
\footnotesize
\movetableright=-20mm
\script{paper_table_tidal.py}
\caption{Flaring tidal star-planet interaction. }
\input{output/table_tidal.tex}
    \label{tab:tidal}
 \tablerefs{\input{output/table_tidal_bibstring.tex}}
\end{table*}

Tidal interaction can also lead to enhanced activity, globally, and locally. Locally, we expect to see periodicity with half the orbital period of the planet, corresponding to the two tidal bulges forming in the stellar envelope~\citep{cuntz2000stellar}. There are neither models to describe how that process leads to enhanced activity, or excess flaring, in star-planet systems, nor are there existing observations. However, in the extreme case of RS CVn binaries, starspots have been observed modulated with half the orbital period of the stellar companion~\citep{ozavci2018recurrent, kriskovics2023ei}. A similar effect was measured when the companion was likely a very low mass star or a brown dwarf~\citep{donati1995activity, frasca2008spots, parks2021interferometric}, and in the case of a K dwarf/white dwarf binary~\citep{hussain2006spot, watson2007roche}. In ellipsoidal binaries, a modulation of flares beyond geometrical effects was observed with Kepler~\citep{gao2016whitelight}, suggesting flaring regions at the location of the extreme tidal bulges in these stars. However, these systems are synchronized, $P_{rot}=P_{orb}$, so that intrinsic active longitudes~\citep{usoskin2007longterm, weber2013theory} cannot be disambiguated from the result of tidal star-star interaction ~\citep{holzwarth2003dynamics}. We can decrease the companion mass further to arrive at local, tidal star-planet interaction, for which, in our sample, spin and orbit are not synchronized. In analogy to~\citet{holzwarth2003dynamics}, we can picture an asymmetry induced in the stellar dynamo by the tidal disturbance, which would lead to a preference for flares at longitudes about $180\,$deg apart from each other.

@@

We applied the same technique as described in Section~\ref{sec:methods:adtest} to determine the significance of a local tidal SPI signal. Following~\citet{ilic2022tidal}, we compare the resulting $p$-values with three different models of tidal interaction as proxies for expected excess flaring due to tides in Fig.~\ref{fig:tidal}. We list the derived values in Table~\ref{tab:tidal}.

The first model measures the torque $\partial L / \partial t$, exerted by angular momentum transfer from star to planet or from planet to star, assumin the star has a convective envelope~(\citealt{penev2012constraining}, their Eqns. 1 and 2). We choose the tidal quality factor $Q_*=10^7$. Our results in Fig~\ref{fig:tidal} (left panel) do not depend on the choice of $Q_*$, as we only need to know the tidal torque down to a proportionality factor. We also make the simplifying assumption that the rotation period of the convective envelope is the same as the surface rotation period measured as described in Section~\ref{sec:data:rotationperiods}, ignoring differential rotation effects.

The second model considers the tidal dissipation timescale -- the shorter the timescale, the higher the power of the interaction~\citep{zahn1977tidal}. Following~\citet{ilic2022tidal}, we adopt Eqn.2 from~\citep{albrecht2012obliquities} for the tidal dissipation timescale in the convective envelope. We assume that all stars in our sample have a convective envelope because they flare, and flaring requires a dynamo-generated magnetic field, which can only operate in stars with convective envelopes. Most of our stars are dwarf stars below the Kraft break at spectral type F5, except for KOI-12, which is an F5 sub-giant~\citep{frasca2016activity}.

The third model uses the gravitational perturbation as a proxy for the interaction~\citep{cuntz2000stellar}, originally proposed for star-planet interactions.  Their Eqn. 1 proposes a gravitational perturbation proportional to the mass ratio between planet and star, and $a^{-3}$.

All models require the planetary and stellar mass to be known which we collect from the literature. For missing planetary masses, radii are known, so we can use the mass-radius relations implemented in \texttt{forecaster}~\citep{chen2017probabilistic}. GJ 3323, GJ 674 and GJ 3082 were also missing stellar mass estimates, which we obtained from the \citep{mann2015how, mann2016erratum} relations between absolute Ks magnitude~\citep{skrutskie2006two} and stellar mass for M dwarfs using the distances from \citet{bailer-jones2021estimating}. We list the masses in Table~\ref{tab:tidal}.
